AiPoker - Implementation of a Poker Agent

### 2014-11-01 report/paper.tex
Fixed some spelling in the report

%% the only exception to this rule is the \firstsection command
\firstsection{Background}
\maketitle 
Texas Hold'em is a popular variant of poker, not just to play with friends but also online against other people for money. The challenge of creating the best poker agent has attracted a lot of people and is a popular subject in AI development.
\\*\\*
\textbf{Type of Games}\\*
There are many different games today and the strategy to solve these games depends on the type of game. In game theory games can be classified due to the knowledge if the transition to a new state within the game depends on a stochastic process and if there is some hidden information. Poker belongs with the most difficult kind of card game to solve discretely since it is an stochastic game with imperfect information\cite{Kleij}.
\\*\\*
\textbf{Perfect Information vs Imperfect Information}\\*
In perfect information games both players can observe the complete state of the game at any given time, this is the case in games such as Chess and Go. Contrary to this we have imperfect information games where each player is unable to observe the complete state, one example for this is poker were neither player can see the other players pocket cards. 
\\*\\*
\textbf{Deterministic vs Stochastic}\\*
In deterministic games the next state of the game is uniquely determined by an action taken at the current state. In contrast, a stochastic game is a game where the player don't have control over the transition to the next state. For example in poker where the new cards is drawn from a scrambled deck.

@@

Davidson et al. (2000)\cite{davidson} tried to improve the opponent modeling in Loki with artificial neural networks (ANNs) to predict the opponent?s actions, and achieves about 85\% accuracy at predicting the moves. This is a significant improvement over the previous implementation and the new system was named Poki. The problems with ANNs is that they can be slow to train and thus they will make weak plays against players who frequently change their play style.

@@

In order to always have money in the pot of each round and take away the risk of a player to be to defensive and never bet a small fee i circulating around the players through the game called the blind. The blind is two amounts called big and small blind. When the game starts a player gets randomly selected to pay the big blind and the player next on the left to him gets to pay the small blind. When a round  is played the blind moves one step to the left so the person who had the small blind get to pay the big blind this turn and the player left to him gets the small blind. If there is only two players they just switch big and small blind every round. The big blind is the double amount of the small blind and is a pre set sum.

@@

\textbf{Pre Flop}\\*
In this stage the players gets their two cards from the dealer called the pocket cards and the player who is first are able to either bet, call if he has not paid the big blind, check if he has paid the big blind or fold. Since you can not form a five combination of cards yet only a speculation of how good your pocket cards are can be made. If a player bets or raises all other players get the opportunity to either call, raise or fold max three times, after that the players who has not folded moves on to the flop.
\\*\\*
\textbf{Flop}\\*
The flop is when the dealer puts out the three first cards on the table. These cards are visible to all players. The pocket cards can now be combined with the flop cards in order to create a strong hand. The players will make moves in the same procedure as in the pre flop stage.
\\*\\*
\textbf{Turn}\\*
Now the dealer adds a fourth card to the flop. The players will now bet again.
\\*\\*
\textbf{River}\\*
The dealer adds a fifth card to the table. This is the last card which is dealt. After the last betting round, the game moves on to showdown. 
\\*\\*
\textbf{Showdown}\\*
In this phase the final betting is completed and the players who has not folded will now show their cards in order to decide who is the winner of this round. The winner will get the full pot and then a new round will start with the pre flop.
\\*\\*
\textbf{The different five combination of cards}\\*
There are nine different types of combination of cards and here they are listed in descending order[3]:
\begin{itemize}
  \item \textit{Straight Flush}: When the player has five cards in a consecutive order of value with the same suit. If more than one player have a Straight Flush the player with the highest values of his/her cards wins. The best variant is the Royal Straight Flush when the player has Ace, King, Queen, Knight and a ten within the same suit.
  \item \textit{Four of a kind}: It is a hand that contains four of the same number, which means in all the different suits. If different players have the same type of hand the hand with the highest four of a kind wins.
  \item \textit{Full House}: This hand is a combination of three of a kind and a pair. If more than one player have a Full House the player with the highest value three of a kind wins, if two players have the same three of a kind the player with the highest value of one pair wins.
  \item \textit{Flush}: A flush is when a player has five cards with the same suit, if more than one player have a flush the one with the highest card wins. If they have the same highest card then the the second highest card is checked and so on.
  \item \textit{Straight}: When you have a five cards in a consecutive order of value. If more than one player has a Straight then the player with the highest card wins.
  \item \textit{Three of a Kind}: When you have three cards of the same value, if more than one player has a Three of a kind then the player with the highest value of the cards wins.
  \item \textit{Two Pair}: When a player has two pair of cards, if more than one player have a two pair the player with the highest pair has the best hand between them.
  \item \textit{One Pair}: When you have one pair of cards, if more than one player have a one pair the player with the highest pair has the best hand between them.
  \item \textit{High Card}: If the card does not match any of the types above then the card with the highest value is considered. 
\end{itemize}

@@

\section{Method}
\subsection{Basic Game Engine}
In order to even begin implementing an AI, the core poker game had to be implemented first. This was made by first implementing an agent that played at random. For simplicity the game implemented in this study is a two player game with the limited edition of Texas Hold?em where a human player faces a computer AI. The player who starts is chosen at random when the game is started. The player that starts then puts one dollar in to the pot. The other player puts two dollars in the pot. Then the game is turn based like ordinary Texas Hold?em, the players can bet, check, call, raise and fold. And after each betting round cards are dealt to the table and the players combine the cards on hand with the cards on the table in order to get the strongest hand possible.

@@

\subsubsection{Hand Strength}
Since Monte Carlo is a heuristic search which simulate a set of outcomes in a game to calculate what action to make the logic so far does not consider how good the cards on the agents hand is compared to other hands. To add this logic to the agent an algorithm could be used that would return a probability of how strong the current hand cards is with the cards on the table. This probability of the Hand Strength may, at the earliest be calculated the the flop has been shown\cite{wenkai}.

The Hand Strength is calculated by comparing the cards on the hand with all possible combination of two cards that the opponent can have and sum the time the agents cards wins compared to the opponent and divide by the number of combinations that was simulated.

@@

\subsubsection{Effective Hand Strength}
By combining the hand strength with the hand potential a probability of winning can be calculated, and this is called the effective hand strength and is calculated as following\cite{wenkai}:

@@

\subsubsection{Bucketing}
Since EHS needs at least the flop in order to be used for decision making, the Bucketing algorithm was implemented  as a complement to the EHS algorithm. This algorithm is only used in in the pre-flop phase where no board cards are visible \cite{schweizer}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.60]{img/bucketlist.png}
    \caption{\label{fig:bucketlist} The bucket list.}
  \end{center}
\end{figure}

@@

\subsubsection{Randomness}
All previously mentioned algorithms take cards as input and gives the probability of the cards in different way to make the best decision with the current hand cards. Since a good poker player needs to be able to bluff, a random function has been implemented in order to give the agent a certain level of unpredictableness. The agent will make a random move approximately 5\% of the time.

@@

\section{Result}
\subsection{The Game}
The result of this study is a Texas Hold'em game with an AI implemented using several different AI techniques and combining them together. The AI in the interactive game is using Monte Carlo Tree Search together with Hand Strength as a heuristic. The game is fully playable with an opponent that is somewhat realistic in the way it chooses what move to make. The 3D-scene is used to visualise the poker table, and it also shows the cards on hand. This is shown in fig(\ref{fig:3dscene}).
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.27]{img/3dscene.png}
    \caption{\label{fig:3dscene} The 3D-scene in the final version showing the player's cards on hand.}
  \end{center}
\end{figure}

@@

\section{Conclusion}
\subsection{Evaluation of the Result}
At a first glance it can look strange that the random ai have so many won hands, but one have to take into consideration that the amount of won hands is not the same as the amount of money won. The random agent wins a lot of hands, but this is due to the other agents folding when they have bad cards. So the average size of the money pot is very small for the hands the random bot wins and much larger for the hands the other agents win.

>>>>

%% the only exception to this rule is the \firstsection command
\firstsection{Background}
\maketitle 
Texas Hold'em is a popular variant of poker, not just to play with friends but also online against other people for money. The challenge of creating the best poker agent has attracted a lot of people and is a popular subject in AI development.
\\*\\*
\textbf{Type of Games}\\*
There are many different games today and the strategy to solve these games depends on the type of game. In game theory games can be classified due to the knowledge if the transition to a new state within the game depends on a stochastic process and if there is some hidden information. Poker belongs with the most difficult kind of card game to solve discretely since it is an stochastic game with imperfect information\cite{Kleij}.
\\*\\*
\textbf{Perfect Information vs Imperfect Information}\\*
In perfect information games both players can observe the complete state of the game at any given time, this is the case in games such as Chess and Go. Contrary to this we have imperfect information games where each player is unable to observe the complete state, one example for this is poker were neither player can see the other players pocket cards. 
\\*\\*
\textbf{Deterministic vs Stochastic}\\*
In deterministic games the next state of the game is uniquely determined by an action taken at the current state. In contrast, a stochastic game is a game where the player does not have control over the transition to the next state. For example in poker where the new cards is drawn from a scrambled deck.

@@

Davidson et al. (2000)\cite{davidson} tried to improve the opponent modeling in Loki with artificial neural networks (ANNs) to predict the opponent's actions, and achieves about 85\% accuracy at predicting the moves. This is a significant improvement over the previous implementation and the new system was named Poki. The problems with ANNs is that they can be slow to train and thus they will make weak plays against players who frequently change their play style.

@@

In order to always have money in the pot of each round and take away the risk of a player to be to defensive and never bet a small fee i circulating around the players through the game called the blind. The blind is two amounts called big and small blind. When the game starts a player gets randomly selected to pay the big blind and the player next on the left to him gets to pay the small blind. When a round is played the blind moves one step to the left so that the person who had the small blind get to pay the big blind this turn and the player left to him gets the small blind. If there are only two players they just switch big and small blind every round. The big blind is the double amount of the small blind and is a pre set sum.

@@

\textbf{Pre Flop}\\*
In this stage the players get two cards each from the dealer. The cards are called the pocket cards and the player who is first is able to either bet, call if he has not paid the big blind, check if he has paid the big blind or fold. Since you can not form a five combination of cards yet only a speculation of how good your pocket cards are can be made. If a player bets or raises all other players get the opportunity to either call, raise or fold max three times, after that the players who has not folded moves on to the flop.
\\*\\*
\textbf{Flop}\\*
The flop is when the dealer puts out the three first cards on the table. These cards are visible to all players. The pocket cards can now be combined with the flop cards in order to create a strong hand. The players will make moves in the same procedure as in the pre flop stage.
\\*\\*
\textbf{Turn}\\*
Now the dealer adds a fourth card to the flop. The players will now bet again.
\\*\\*
\textbf{River}\\*
The dealer adds a fifth card to the table. This is the last card which is dealt. After the last betting round, the game moves on to showdown. 
\\*\\*
\textbf{Showdown}\\*
In this phase the final betting is completed and the players who has not folded will now show their cards in order to decide who is the winner of this round. The winner will get the full pot and then a new round will start with the pre flop.
\\*\\*
\textbf{The different five combination of cards}\\*
There are nine different types of combination of cards and here they are listed in descending order:
\begin{itemize}
  \item \textit{Straight Flush}: When the player has five cards in a consecutive order of value within the same suit. If more than one player have a Straight Flush the player with the highest values of his/her cards wins. The best variant is the Royal Straight Flush when the player has Ace, King, Queen, Knight and a ten within the same suit.
  \item \textit{Four of a kind}: It is a hand that contains four of the same value, which means in all the different suits. If different players have the same type of hand the hand with the highest four of a kind wins.
  \item \textit{Full House}: This hand is a combination of three of a kind and a pair. If more than one player have a Full House the player with the highest value three of a kind wins, if two players have the same three of a kind the player with the highest value of one pair wins.
  \item \textit{Flush}: A flush is when a player has five cards with the same suit, if more than one player have a flush the one with the highest card wins. If they have the same highest card then the the second highest card is checked and so on.
  \item \textit{Straight}: When you have a five cards in a consecutive order of value. If more than one player has a Straight then the player with the highest card wins.
  \item \textit{Three of a Kind}: When you have three cards of the same value, if more than one player has a Three of a kind then the player with the highest value of the cards wins.
  \item \textit{Two Pair}: When a player has two pair of cards, if more than one player have a two pair the player with the highest pair has the best hand between them.
  \item \textit{One Pair}: When you have one pair of cards, if more than one player have a one pair the player with the highest pair has the best hand between them.
  \item \textit{High Card}: If the card does not match any of the types above then the card with the highest value is considered. 
\end{itemize}

@@

\section{Method}
\subsection{Basic Game Engine}
In order to even begin implementing an AI, the core poker game had to be implemented first. This was made by first implementing an agent that played at random. For simplicity the game implemented in this study is a two player game with the limited edition of Texas Hold'em where a human player faces a computer AI. The player who starts is chosen at random when the game is started. The player that starts then puts one dollar in to the pot. The other player puts two dollars in the pot. Then the game is turn based like ordinary Texas Hold'em, the players can bet, check, call, raise and fold. And after each betting round cards are dealt to the table and the players combine the cards on hand with the cards on the table in order to get the strongest hand possible.

@@

\subsubsection{Hand Strength}
Since Monte Carlo is a heuristic search which simulate a set of outcomes in a game to calculate what action to make the logic so far does not consider how good the cards on the agents hand is compared to other hands. To add this logic to the agent an algorithm could be used that would return a probability of how strong the current hand cards is with the cards on the table. This probability of the Hand Strength may, at the earliest be calculated when the flop has been dealt\cite{wenkai}.

The Hand Strength is calculated by comparing the cards on the hand with all possible combination of two cards that the opponent can have and sum the times the agent's cards wins compared to the opponent and divide by the number of combinations that was simulated.

@@

\subsubsection{Effective Hand Strength}
By combining the hand strength with the hand potential a probability of winning can be calculated, and this is called the effective hand strength and is calculated as follows\cite{wenkai}:

@@

\subsubsection{Bucketing}
Since effective hand strength (EHS) needs at least the flop in order to be used for decision making, the Bucketing algorithm was implemented  as a complement to the EHS algorithm. This algorithm is only used in in the pre-flop phase where no board cards are visible \cite{schweizer}.
\begin{figure}[here]
  \begin{center}
    \includegraphics[scale=0.60]{img/bucketlist.png}
    \caption{\label{fig:bucketlist} The bucket list.}
  \end{center}
\end{figure}

@@

\subsubsection{Randomness}
All previously mentioned algorithms take cards as input and gives the probability of the cards in different way to make the best decision with the current hand cards. Since a good poker player needs to be able to bluff, a random function has been implemented in order to give the agent a certain level of unpredictability. The agent will make a random move approximately 5\% of the time.

@@

\section{Result}
\subsection{The Game}
The result of this study is a Texas Hold'em game with an AI implemented using several different AI techniques and combining them together. The AI in the interactive game is using Monte Carlo Tree Search together with Hand Strength as a heuristic. The game is fully playable with an opponent that is somewhat realistic in the way it chooses what move to make. The 3D-scene is used to visualize the poker table, and it also shows the cards on hand. This is shown in fig(\ref{fig:3dscene}).
\begin{figure}[H]
  \begin{center}
    \includegraphics[scale=0.27]{img/3dscene.png}
    \caption{\label{fig:3dscene} The 3D-scene in the final version showing the player's cards on hand.}
  \end{center}
\end{figure}

@@

\section{Conclusion}
\subsection{Evaluation of the Result}
At a first glance it can look strange that the random agent won so many hands, but one have to take into consideration that the amount of won hands is not the same as the amount of money won. The random agent wins a lot of hands, but this is due to the other agents folding when they have bad cards. So the average size of the money pot is very small for the hands the random bot wins and much larger for the hands the other agents win.
